---
title: "Adriatic Sea"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---
```{r library}
#| cache: false
#| echo: false
#| warning: false

library(tidyverse)
library(lubridate)
library(brms)
library(purrr)
library(dplyr)
library(Utilities.Package)
```

```{r Open Chlorophyll}
#| cache: false
#| echo: false
#| warning: false

# Read in
df_chla <- read.csv("Data/Variables_values/Monthly_Chl_boxes.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  pivot_longer(-Date, names_to = "metric", values_to = "values") %>% 
  mutate(Box = gsub(".*_","",metric),
         Metric = case_when(str_detect(metric,"GEOM") ~ "GeoMean_Chla",
                            T ~ "STD_Chla")) %>% 
  select(-metric) %>% 
  pivot_wider(names_from = Metric, values_from = values) %>% 
  mutate(Box = case_when(Box == "Adriatico" ~ "Adriatic",
                         Box == "Chipre" ~ "Cyprus",
                         T ~ "Lyon"))


```

```{r Open HF}
#| cache: false
#| echo: false
#| warning: false

df_HF_Adriatic <- read.csv("Data/Variables_values/Monthly_HF_Adriatic.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_HF_Lyon <- read.csv("Data/Variables_values/Monthly_HF_Lyon.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")

df_HF_Chipre <- read.csv("Data/Variables_values/Monthly_HF_Cyprus.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_HF <- rbind(df_HF_Adriatic,df_HF_Lyon,df_HF_Chipre)

rm(df_HF_Adriatic)
rm(df_HF_Lyon)
rm(df_HF_Chipre)


```

```{r Open MLD}
#| cache: false
#| echo: false
#| warning: false

df_MLD_Adriatic <- read.csv("Data/Variables_values/Monthly_MLD_Adriatic.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_MLD_Lyon <- read.csv("Data/Variables_values/Monthly_MLD_Lyon.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")

df_MLD_Chipre <- read.csv("Data/Variables_values/Monthly_MLD_Cyprus.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_MLD <- rbind(df_MLD_Adriatic,df_MLD_Lyon,df_MLD_Chipre)

rm(df_MLD_Adriatic)
rm(df_MLD_Lyon)
rm(df_MLD_Chipre)

```

```{r Open NH4}
#| cache: false
#| echo: false
#| warning: false


df_NH4_Adriatic <- read.csv("Data/Variables_values/Monthly_NH4_Adriatic_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_NH4_Lyon <- read.csv("Data/Variables_values/Monthly_NH4_Lyon_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")

df_NH4_Chipre <- read.csv("Data/Variables_values/Monthly_NH4_Cyprus_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_NH4 <- rbind(df_NH4_Adriatic,df_NH4_Lyon,df_NH4_Chipre)

rm(df_NH4_Adriatic)
rm(df_NH4_Lyon)
rm(df_NH4_Chipre)

```

```{r Open NO3}
#| cache: false
#| echo: false
#| warning: false


df_NO3_Adriatic <- read.csv("Data/Variables_values/Monthly_NO3_Adriatic_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_NO3_Lyon <- read.csv("Data/Variables_values/Monthly_NO3_Lyon_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")

df_NO3_Chipre <- read.csv("Data/Variables_values/Monthly_NO3_Cyprus_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_NO3 <- rbind(df_NO3_Adriatic,df_NO3_Lyon,df_NO3_Chipre)

rm(df_NO3_Adriatic)
rm(df_NO3_Lyon)
rm(df_NO3_Chipre)

```

```{r Open PO4}
#| cache: false
#| echo: false
#| warning: false

df_PO4_Adriatic <- read.csv("Data/Variables_values/Monthly_PO4_Adriatic_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_PO4_Lyon <- read.csv("Data/Variables_values/Monthly_PO4_Lyon_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")

df_PO4_Chipre <- read.csv("Data/Variables_values/Monthly_PO4_Cyprus_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_PO4 <- rbind(df_PO4_Adriatic,df_PO4_Lyon,df_PO4_Chipre)

rm(df_PO4_Adriatic)
rm(df_PO4_Lyon)
rm(df_PO4_Chipre)

```

```{r Open SSS}
#| cache: false
#| echo: false
#| warning: false

df_SSS_Adriatic <- read.csv("Data/Variables_values/Monthly_SSS_Adriatic.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_SSS_Lyon <- read.csv("Data/Variables_values/Monthly_SSS_Lyon.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")

df_SSS_Chipre <- read.csv("Data/Variables_values/Monthly_SSS_Cyprus.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_SSS <- rbind(df_SSS_Adriatic,df_SSS_Lyon,df_SSS_Chipre)

rm(df_SSS_Adriatic)
rm(df_SSS_Lyon)
rm(df_SSS_Chipre)

```

```{r Open SST}
#| cache: false
#| echo: false
#| warning: false

df_SST <- read.csv("Data/Variables_values/Monthly_SST_boxes.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  pivot_longer(-Date, names_to = "metric", values_to = "values") %>% 
  mutate(Box = gsub(".*_","",metric),
         Metric = "SST") %>% 
  select(-metric) %>% 
  pivot_wider(names_from = Metric, values_from = values) %>% 
  mutate(Box = case_when(Box == "Adriatico" ~ "Adriatic",
                         Box == "Chipre" ~ "Cyprus",
                         T ~ "Lyon"))
```

```{r stack all}
#| cache: false
#| echo: false
#| warning: false

date_min <- dmy("01-01-2018")
date_max <- dmy("01-01-2021")

df_chla <- df_chla %>% 
  dplyr::filter(Date >= min(df_HF$Date), 
                Date <= max(df_HF$Date))

df <- df_chla %>%
  left_join(df_SST,  by = c("Box", "Date")) %>%
  left_join(df_SSS,  by = c("Box", "Date")) %>%
  left_join(df_PO4,  by = c("Box", "Date")) %>%
  left_join(df_NO3,  by = c("Box", "Date")) %>%
  left_join(df_NH4,  by = c("Box", "Date")) %>%
  left_join(df_MLD,  by = c("Box", "Date")) %>%
  left_join(df_HF,   by = c("Box", "Date")) %>% 
  pivot_longer(-c(Date,Box), names_to = "Metric", values_to = "Value")


write.csv(df,"Data/df.all.csv", row.names = F)

# df %>% 
#   # dplyr::filter(Date >= date_min, 
#                 # Date <= date_max) %>% 
#   ggplot(aes(x = Date, y = Value, color = Box, group = Box))+
#   geom_point()+
#   geom_line()+
#   # geom_smooth(method = "gam")+
#   facet_wrap(~Metric,scales = "free")+
#   theme_bw()

df_wide <- df %>% 
  pivot_wider(names_from = Metric,values_from = Value)

# pairs(df_wide %>% dplyr::select(-Box))


```

```{r Model Adriatic}
#| cache: false
#| echo: false
#| warning: false

library(brms)
library(tidyverse)
library(patchwork)

df_wide_Adriatic <- df_wide %>%
  dplyr::filter(Box == "Adriatic") %>%
  mutate(
    DIN  = NO3 + NH4,
    N_P  = DIN / PO4,
    # log_N_P = log1p(N_P),
    sum_Nutri = NO3 + NH4 + PO4
  ) %>%
  dplyr::select(-c(PO4,NO3,NH4,DIN))

df_wide_Adriatic_STD <- df_wide_Adriatic %>%
  mutate(across(c(SST, SSS, N_P, sum_Nutri, MLD, HF),
                ~ ( .x - mean(.x, na.rm = TRUE) ) / sd(.x, na.rm = TRUE )))

# pairs(df_wide_Adriatic_STD %>% dplyr::select(-Box))

df_back_to_raw_Adriatic <- df_wide_Adriatic %>% 
  pivot_longer(-c(Date, Box, GeoMean_Chla, STD_Chla), values_to = "values", names_to = "metrics") %>% 
  group_by(metrics) %>% 
  reframe(avg = mean(values, na.rm = T),
          sd = sd(values, na.rm=T))


fit_Adriatic <- brm(
  formula = GeoMean_Chla ~
    s(SST) + s(SSS) + s(N_P) +
    s(sum_Nutri) + s(MLD) +
    s(HF),
  data = df_wide_Adriatic_STD,
  family = Gamma(link = "log"),
  chains = 4, cores = 15,
  iter = 10000,
  warmup = 1000,
  control = list(adapt_delta = 0.97,  max_treedepth = 15),
  file = "Data/Model/Chla_GAM_all_10000_Adriatic_log_adapt_delta_97_tree_15_Cleaned_STD"
)

# plot(fit_adriatic)
# summary(fit_Adriatic)




### Plots Adriatic

SST_back_to_raw_Adriatic <- df_back_to_raw_Adriatic %>% 
  dplyr::filter(metrics == "SST")

Adriatic_sst <- conditional_effects(
  fit_Adriatic,
  effects = "SST",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]] %>% 
ggplot(aes(x = (SST*SST_back_to_raw_Adriatic$sd)+SST_back_to_raw_Adriatic$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "darkred", alpha = 0.1)+
  geom_line(color= "darkred")+
  labs(y = "Chla (mg.m⁻³)",x="SST (°C)")+
  theme_Bede()


SSS_back_to_raw_Adriatic <- df_back_to_raw_Adriatic %>% 
  dplyr::filter(metrics == "SSS")

Adriatic_sss <- conditional_effects(
  fit_Adriatic,
  effects = "SSS",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]]%>% 
ggplot(aes(x = (SSS*SSS_back_to_raw_Adriatic$sd)+SSS_back_to_raw_Adriatic$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "#62B0D1", alpha = 0.1)+
  geom_line(color= "#62B0D1")+
  labs(y = "Chla (mg.m⁻³)",x="SSS (ppm)")+
  theme_Bede()


N_P_back_to_raw_Adriatic <- df_back_to_raw_Adriatic %>% 
  dplyr::filter(metrics == "N_P")

Adriatic_N_P <- conditional_effects(
  fit_Adriatic,
  effects = "N_P",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]]%>% 
ggplot(aes(x = (N_P*N_P_back_to_raw_Adriatic$sd)+N_P_back_to_raw_Adriatic$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "yellow3", alpha = 0.1)+
  geom_line(color= "yellow3")+
  labs(y = "Chla (mg.m⁻³)",x="NPratio")+
  theme_Bede()






sum_Nutri_back_to_raw_Adriatic <- df_back_to_raw_Adriatic %>% 
  dplyr::filter(metrics == "sum_Nutri")

Adriatic_sum_Nutri <- conditional_effects(
  fit_Adriatic,
  effects = "sum_Nutri",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]]%>% 
ggplot(aes(x = (sum_Nutri*sum_Nutri_back_to_raw_Adriatic$sd)+sum_Nutri_back_to_raw_Adriatic$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "darkgreen", alpha = 0.1)+
  geom_line(color= "darkgreen")+
  labs(y = "Chla (mg.m⁻³)",x="Total Nutrients (mmol.m-3)")+
  theme_Bede()



MLD_back_to_raw_Adriatic <- df_back_to_raw_Adriatic %>% 
  dplyr::filter(metrics == "MLD")

Adriatic_MLD <- conditional_effects(
  fit_Adriatic,
  effects = "MLD",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]]%>% 
ggplot(aes(x = (MLD*MLD_back_to_raw_Adriatic$sd)+MLD_back_to_raw_Adriatic$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "darkblue", alpha = 0.1)+
  geom_line(color= "darkblue")+
  labs(y = "Chla (mg.m⁻³)",x="Mixed Layer Depth (m)")+
  theme_Bede()


HF_back_to_raw_Adriatic <- df_back_to_raw_Adriatic %>% 
  dplyr::filter(metrics == "HF")

Adriatic_HF <- conditional_effects(
  fit_Adriatic,
  effects = "HF",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]]%>% 
ggplot(aes(x = (HF*HF_back_to_raw_Adriatic$sd)+HF_back_to_raw_Adriatic$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "orangered", alpha = 0.1)+
  geom_line(color= "orangered")+
  labs(y = "Chla (mg.m⁻³)",x="Heat Flux (W.m-2)")+
  theme_Bede()



plot_Adriatic <- (Adriatic_sst+Adriatic_HF)/
 ( Adriatic_N_P+Adriatic_sum_Nutri)+
  (Adriatic_MLD+Adriatic_sss)




# plot(conditional_smooths(fit_adriatic), ask = FALSE)

```

```{r Flextable Adriatic Model}
#| label: tbl-Model_outputs_Adriatic
#| fig-cap: Posterior summary of fixed‐effect and smooth‐term estimates from the generalized additive model (Gamma family, log link) predicting GeoMean_Chla as a function of sea surface temperature (SST), sea surface salinity (SSS), N:P ratio, total nutrient load (sum_Nutri), mixed‐layer depth (MLD), and heat flux (HF). Estimates shown are posterior means (Estimate) ± posterior standard errors (Est. Error), 95 % credible intervals (l–95 % CI, u–95 % CI), Gelman–Rubin convergence diagnostic (Rhat), and effective sample sizes for bulk and tail (Bulk ESS, Tail ESS).
#| echo: false
#| error: false
#| message: false
#| warning: false
#| out-width: "95%"
#–– load packages
library(brms)
library(dplyr)
library(tibble)
library(flextable)

#–– extract summary lists
s <- summary(fit_Adriatic)

# Population-level (fixed) effects
fixed_df <- as.data.frame(s$fixed) %>%
  rownames_to_column("Parameter") %>%
  mutate(Term = "Fixed")

# Smooth terms
smooth_df <- as.data.frame(s$smooth) %>%
  rownames_to_column("Parameter") %>%
  mutate(Term = "Smooth")

# Family‐specific parameters (e.g. Gamma shape)
spec_df <- as.data.frame(s$spec_pars) %>%
  rownames_to_column("Parameter") %>%
  mutate(Term = "Family")

# bind into one table
res_df <- bind_rows(fixed_df, smooth_df, spec_df) %>%
  select(Term, Parameter, Estimate = Estimate,
         `Est.Error`, `l-95% CI`, `u-95% CI`, Rhat, `Bulk_ESS`, `Tail_ESS`)

#–– build flextable
ft <- flextable(res_df) %>%
  # format numeric columns to 3 decimal places
  colformat_num(j = c("Estimate","Est.Error","l-95% CI","u-95% CI","Rhat","Bulk_ESS","Tail_ESS"),
                digits = 3) %>%
  # add a little zebra striping and autofit
  theme_zebra() %>%
  autofit() %>%
  # optional: bold the header row
  bold(i = 1, bold = TRUE)

# display
ft

```



@tbl-Model_outputs_Adriatic shows that Sea‐surface temperature emerged as the only clearly significant driver of chlorophyll‐a, with the smooth term estimated at –2.72 (95 % CI –5.32 to –0.72), indicating a marked decline in GeoMean_Chl a as temperatures rise. The N : P ratio showed a moderate negative estimate of –1.71, but its 95 % CI (–3.96 to 0.07) just overlaps zero, suggesting a weak downward trend. In contrast, sea‐surface salinity (–0.40; 95 % CI –1.05 to 0.40), total nutrient load (2.65; 95 % CI –2.69 to 8.52), mixed‐layer depth (–0.67; 95 % CI –4.17 to 3.63), and surface heat flux (0.21; 95 % CI –1.14 to 1.72) all produced credible intervals spanning zero, indicating no detectable effect on chlorophyll‐a in this model. The Gamma‐family shape parameter was estimated at 34.84 (95 % CI 29.01 to 41.14).

```{r overall raw plot}
#| label: fig-Adriatic_Plots
#| fig-cap: "Posterior conditional (marginal) effects of the six environmental drivers retained in the Gamma-GAM for monthly geometric-mean chlorophyll-a (Chla) in the Gulf of Lyon. For each predictor, the solid coloured line is the posterior median of the expected Chla when that variable varies across its observed range while all other covariates are fixed at their means; the shaded band is the corresponding 95 % credible interval on the response scale. Axes are shown in original physical units: sea-surface temperature (°C), net heat flux (W m⁻²), molar N : P ratio, summed dissolved nutrients (mmol m⁻³), mixed-layer depth (m) and sea-surface salinity (ppt)."
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig-height: 10

plot_Adriatic
```

@fig-Adriatic_Plots displays the posterior mean smooths (solid lines) and their 95 % credible ribbons for each of the six covariates, expressed on the original physical scales. 
As sea‐surface temperature increases from approximately 15 °C to 27 °C, modeled Chl a falls steeply from about 0.16 mg m⁻³ to 0.07 mg m⁻³, with the greatest sensitivity between 15 and 20 °C before the curve asymptotes at warmer temperatures. Surface heat flux produces a unimodal response: Chl a is lowest (~0.065 mg m⁻³) under strong cooling (–300 W m⁻²), rises to a maximum of roughly 0.09 mg m⁻³ at around –50 W m⁻², then declines again to near 0.085 mg m⁻³ by +200 W m⁻². The N : P ratio effect is characterized by an abrupt drop from about 0.12 mg m⁻³ at low ratios (~20) to 0.08 mg m⁻³ by an N : P of 100, after which Chl a plateaus to approximately 0.09 mg m⁻³ at the highest ratios (~100 to 400). Total nutrient concentration drives a pronounced hump: Chl a increases from ~0.06 mg m⁻³ at minimal loads to ~0.10 mg m⁻³ by 100 mmol m⁻³, peaks near 0.18 mg m⁻³ at 500 mmol m⁻³, and then falls back toward 0.10 mg m⁻³ at 725 mmol m⁻³. Mixed‐layer depth exhibits a U‐shaped trough, with Chl a declining from ~0.11 mg m⁻³ at shallow depths (10 m) to ~0.045 mg m⁻³ by ~150 m before climbing slightly to ~0.05 mg m⁻³ at 500 m. Finally, sea‐surface salinity shows a gentle negative trend, with Chl a decreasing from roughly 0.09 mg m⁻³ at 38.25 ppm to about 0.07 mg m⁻³ at 39.2 ppm.