---
title: "Cyprus"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r library}
#| cache: false
#| echo: false
#| warning: false

library(tidyverse)
library(lubridate)
library(brms)
library(purrr)
library(dplyr)
library(Utilities.Package)
```

```{r Open Chlorophyll}
#| cache: false
#| echo: false
#| warning: false

# Read in
df_chla <- read.csv("Data/Variables_values/Monthly_Chl_boxes.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  pivot_longer(-Date, names_to = "metric", values_to = "values") %>% 
  mutate(Box = gsub(".*_","",metric),
         Metric = case_when(str_detect(metric,"GEOM") ~ "GeoMean_Chla",
                            T ~ "STD_Chla")) %>% 
  select(-metric) %>% 
  pivot_wider(names_from = Metric, values_from = values) %>% 
  mutate(Box = case_when(Box == "Adriatico" ~ "Adriatic",
                         Box == "Chipre" ~ "Cyprus",
                         T ~ "Lyon"))


```

```{r Open HF}
#| cache: false
#| echo: false
#| warning: false

df_HF_Adriatic <- read.csv("Data/Variables_values/Monthly_HF_Adriatic.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_HF_Lyon <- read.csv("Data/Variables_values/Monthly_HF_Lyon.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")

df_HF_Chipre <- read.csv("Data/Variables_values/Monthly_HF_Cyprus.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_HF <- rbind(df_HF_Adriatic,df_HF_Lyon,df_HF_Chipre)

rm(df_HF_Adriatic)
rm(df_HF_Lyon)
rm(df_HF_Chipre)


```

```{r Open MLD}
#| cache: false
#| echo: false
#| warning: false

df_MLD_Adriatic <- read.csv("Data/Variables_values/Monthly_MLD_Adriatic.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_MLD_Lyon <- read.csv("Data/Variables_values/Monthly_MLD_Lyon.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")%>% 
  rename(MLD="MLD_mean")

df_MLD_Chipre <- read.csv("Data/Variables_values/Monthly_MLD_Cyprus.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_MLD <- rbind(df_MLD_Adriatic,df_MLD_Lyon,df_MLD_Chipre)

rm(df_MLD_Adriatic)
rm(df_MLD_Lyon)
rm(df_MLD_Chipre)

```

```{r Open NH4}
#| cache: false
#| echo: false
#| warning: false


df_NH4_Adriatic <- read.csv("Data/Variables_values/Monthly_NH4_Adriatic_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_NH4_Lyon <- read.csv("Data/Variables_values/Monthly_NH4_Lyon_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")

df_NH4_Chipre <- read.csv("Data/Variables_values/Monthly_NH4_Cyprus_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_NH4 <- rbind(df_NH4_Adriatic,df_NH4_Lyon,df_NH4_Chipre)

rm(df_NH4_Adriatic)
rm(df_NH4_Lyon)
rm(df_NH4_Chipre)

```

```{r Open NO3}
#| cache: false
#| echo: false
#| warning: false


df_NO3_Adriatic <- read.csv("Data/Variables_values/Monthly_NO3_Adriatic_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_NO3_Lyon <- read.csv("Data/Variables_values/Monthly_NO3_Lyon_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")

df_NO3_Chipre <- read.csv("Data/Variables_values/Monthly_NO3_Cyprus_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_NO3 <- rbind(df_NO3_Adriatic,df_NO3_Lyon,df_NO3_Chipre)

rm(df_NO3_Adriatic)
rm(df_NO3_Lyon)
rm(df_NO3_Chipre)

```

```{r Open PO4}
#| cache: false
#| echo: false
#| warning: false

df_PO4_Adriatic <- read.csv("Data/Variables_values/Monthly_PO4_Adriatic_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_PO4_Lyon <- read.csv("Data/Variables_values/Monthly_PO4_Lyon_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")

df_PO4_Chipre <- read.csv("Data/Variables_values/Monthly_PO4_Cyprus_for_MLD.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_PO4 <- rbind(df_PO4_Adriatic,df_PO4_Lyon,df_PO4_Chipre)

rm(df_PO4_Adriatic)
rm(df_PO4_Lyon)
rm(df_PO4_Chipre)

```

```{r Open SSS}
#| cache: false
#| echo: false
#| warning: false

df_SSS_Adriatic <- read.csv("Data/Variables_values/Monthly_SSS_Adriatic.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Adriatic")

df_SSS_Lyon <- read.csv("Data/Variables_values/Monthly_SSS_Lyon.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Lyon")

df_SSS_Chipre <- read.csv("Data/Variables_values/Monthly_SSS_Cyprus.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  mutate(Box = "Cyprus")

df_SSS <- rbind(df_SSS_Adriatic,df_SSS_Lyon,df_SSS_Chipre)

rm(df_SSS_Adriatic)
rm(df_SSS_Lyon)
rm(df_SSS_Chipre)

```

```{r Open SST}
#| cache: false
#| echo: false
#| warning: false

df_SST <- read.csv("Data/Variables_values/Monthly_SST_boxes.csv", stringsAsFactors = FALSE) %>% 
  mutate(Date = dmy(Date)) %>% 
  as_tibble() %>% 
  pivot_longer(-Date, names_to = "metric", values_to = "values") %>% 
  mutate(Box = gsub(".*_","",metric),
         Metric = "SST") %>% 
  select(-metric) %>% 
  pivot_wider(names_from = Metric, values_from = values) %>% 
  mutate(Box = case_when(Box == "Adriatico" ~ "Adriatic",
                         Box == "Chipre" ~ "Cyprus",
                         T ~ "Lyon"))
```

```{r stack all}
#| cache: false
#| echo: false
#| warning: false

date_min <- dmy("01-01-2018")
date_max <- dmy("01-01-2021")

df_chla <- df_chla %>% 
  dplyr::filter(Date >= min(df_HF$Date), 
                Date <= max(df_HF$Date))

df <- df_chla %>%
  left_join(df_SST,  by = c("Box", "Date")) %>%
  left_join(df_SSS,  by = c("Box", "Date")) %>%
  left_join(df_PO4,  by = c("Box", "Date")) %>%
  left_join(df_NO3,  by = c("Box", "Date")) %>%
  left_join(df_NH4,  by = c("Box", "Date")) %>%
  left_join(df_MLD,  by = c("Box", "Date")) %>%
  left_join(df_HF,   by = c("Box", "Date")) %>% 
  pivot_longer(-c(Date,Box), names_to = "Metric", values_to = "Value") %>% 
  group_by(Metric, Box) %>% 
  mutate(Value = scales::squish(Value, c(stats::quantile(Value, probs = 0.01),stats::quantile(Value, probs = 0.99))))


write.csv(df,"Data/df.all.csv", row.names = F)

# df %>% 
#   # dplyr::filter(Date >= date_min, 
#                 # Date <= date_max) %>% 
#   ggplot(aes(x = Date, y = Value, color = Box, group = Box))+
#   geom_point()+
#   geom_line()+
#   # geom_smooth(method = "gam")+
#   facet_wrap(~Metric,scales = "free")+
#   theme_bw()

df_wide <- df %>% 
  pivot_wider(names_from = Metric,values_from = Value)

# pairs(df_wide %>% dplyr::select(-Box))


```

```{r Model Cyprus}
#| cache: false
#| echo: false
#| warning: false

library(brms)
library(tidyverse)
library(patchwork)

df_wide_Cyprus <- df_wide %>%
  dplyr::filter(Box == "Cyprus") %>%
  mutate(
    DIN  = NO3 + NH4,
    N_P  = DIN / PO4,
    # log_N_P = log1p(N_P),
    sum_Nutri = NO3 + NH4 + PO4
  ) %>%
  dplyr::select(-c(PO4,NO3,NH4,DIN))

df_wide_Cyprus_STD <- df_wide_Cyprus %>%
  mutate(across(c(SST, SSS, N_P, sum_Nutri, MLD, HF),
                ~ ( .x - mean(.x, na.rm = TRUE) ) / sd(.x, na.rm = TRUE )))

# pairs(df_wide_Cyprus_STD %>% dplyr::select(-Box))

df_back_to_raw_Cyprus <- df_wide_Cyprus %>% 
  pivot_longer(-c(Date, Box, GeoMean_Chla, STD_Chla), values_to = "values", names_to = "metrics") %>% 
  group_by(metrics) %>% 
  reframe(avg = mean(values, na.rm = T),
          sd = sd(values, na.rm=T))


fit_Cyprus <- brm(
  formula = GeoMean_Chla ~
    s(SST) + s(SSS) + s(N_P) +
    s(sum_Nutri) + s(MLD) +
    s(HF),
  data = df_wide_Cyprus_STD,
  family = Gamma(link = "log"),
  chains = 4, cores = 15,
  iter = 10000,
  warmup = 1000,
  control = list(adapt_delta = 0.97,  max_treedepth = 15),
  file = "Data/Model/Chla_GAM_all_10000_Cyprus_log_adapt_delta_97_tree_15_Cleaned_STD2"
)

# plot(fit_adriatic)
# summary(fit_Cyprus)




### Plots Cyprus

SST_back_to_raw_Cyprus <- df_back_to_raw_Cyprus %>% 
  dplyr::filter(metrics == "SST")

Cyprus_sst <- conditional_effects(
  fit_Cyprus,
  effects = "SST",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]] %>% 
ggplot(aes(x = (SST*SST_back_to_raw_Cyprus$sd)+SST_back_to_raw_Cyprus$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "darkred", alpha = 0.1)+
  geom_line(color= "darkred")+
  labs(y = "Chla (mg.m⁻³)",x="SST (°C)")+
  theme_Bede()


SSS_back_to_raw_Cyprus <- df_back_to_raw_Cyprus %>% 
  dplyr::filter(metrics == "SSS")

Cyprus_sss <- conditional_effects(
  fit_Cyprus,
  effects = "SSS",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]]%>% 
ggplot(aes(x = (SSS*SSS_back_to_raw_Cyprus$sd)+SSS_back_to_raw_Cyprus$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "#62B0D1", alpha = 0.1)+
  geom_line(color= "#62B0D1")+
  labs(y = "Chla (mg.m⁻³)",x="SSS (ppm)")+
  theme_Bede()


N_P_back_to_raw_Cyprus <- df_back_to_raw_Cyprus %>% 
  dplyr::filter(metrics == "N_P")

Cyprus_N_P <- conditional_effects(
  fit_Cyprus,
  effects = "N_P",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]]%>% 
ggplot(aes(x = (N_P*N_P_back_to_raw_Cyprus$sd)+N_P_back_to_raw_Cyprus$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "yellow3", alpha = 0.1)+
  geom_line(color= "yellow3")+
  labs(y = "Chla (mg.m⁻³)",x="NPratio")+
  theme_Bede()






sum_Nutri_back_to_raw_Cyprus <- df_back_to_raw_Cyprus %>% 
  dplyr::filter(metrics == "sum_Nutri")

Cyprus_sum_Nutri <- conditional_effects(
  fit_Cyprus,
  effects = "sum_Nutri",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]]%>% 
ggplot(aes(x = (sum_Nutri*sum_Nutri_back_to_raw_Cyprus$sd)+sum_Nutri_back_to_raw_Cyprus$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "darkgreen", alpha = 0.1)+
  geom_line(color= "darkgreen")+
  labs(y = "Chla (mg.m⁻³)",x="Total Nutrients (mmol m-3)")+
  theme_Bede()



MLD_back_to_raw_Cyprus <- df_back_to_raw_Cyprus %>% 
  dplyr::filter(metrics == "MLD")

Cyprus_MLD <- conditional_effects(
  fit_Cyprus,
  effects = "MLD",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]]%>% 
ggplot(aes(x = (MLD*MLD_back_to_raw_Cyprus$sd)+MLD_back_to_raw_Cyprus$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "darkblue", alpha = 0.1)+
  geom_line(color= "darkblue")+
  labs(y = "Chla (mg.m⁻³)",x="Mixed Layer Depth (m)")+
  theme_Bede()


HF_back_to_raw_Cyprus <- df_back_to_raw_Cyprus %>% 
  dplyr::filter(metrics == "HF")

Cyprus_HF <- conditional_effects(
  fit_Cyprus,
  effects = "HF",      
  prob    = 0.80        
  # re_formula = NA     
)[[1]]%>% 
ggplot(aes(x = (HF*HF_back_to_raw_Cyprus$sd)+HF_back_to_raw_Cyprus$avg, y = estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), color = NA, fill = "orangered", alpha = 0.1)+
  geom_line(color= "orangered")+
  labs(y = "Chla (mg.m⁻³)",x="Heat Flux (W.m-2)")+
  theme_Bede()



plot_Cyprus <- (Cyprus_sst+Cyprus_HF)/
 ( Cyprus_N_P+Cyprus_sum_Nutri)+
  (Cyprus_MLD+Cyprus_sss)




# plot(conditional_smooths(fit_adriatic), ask = FALSE)

```

```{r Flextable Cyprus Model}
#| label: tbl-Model_outputs_Cyprus
#| fig-cap: Posterior summary of fixed‐effect and smooth‐term estimates from the generalized additive model (Gamma family, log link) predicting GeoMean_Chla as a function of sea surface temperature (SST), sea surface salinity (SSS), N:P ratio, total nutrient load (sum_Nutri), mixed‐layer depth (MLD), and heat flux (HF). Estimates shown are posterior means (Estimate) ± posterior standard errors (Est. Error), 95 % credible intervals (l–95 % CI, u–95 % CI), Gelman–Rubin convergence diagnostic (Rhat), and effective sample sizes for bulk and tail (Bulk ESS, Tail ESS).
#| echo: false
#| error: false
#| message: false
#| warning: false
#| out-width: "95%"
#–– load packages
library(brms)
library(dplyr)
library(tibble)
library(flextable)

#–– extract summary lists
s <- summary(fit_Cyprus)

# Population-level (fixed) effects
fixed_df <- as.data.frame(s$fixed) %>%
  rownames_to_column("Parameter") %>%
  mutate(Term = "Fixed")

# Smooth terms
smooth_df <- as.data.frame(s$smooth) %>%
  rownames_to_column("Parameter") %>%
  mutate(Term = "Smooth")

# Family‐specific parameters (e.g. Gamma shape)
spec_df <- as.data.frame(s$spec_pars) %>%
  rownames_to_column("Parameter") %>%
  mutate(Term = "Family")

# bind into one table
res_df <- bind_rows(fixed_df, smooth_df, spec_df) %>%
  select(Term, Parameter, Estimate = Estimate,
         `Est.Error`, `l-95% CI`, `u-95% CI`, Rhat, `Bulk_ESS`, `Tail_ESS`)

#–– build flextable
ft <- flextable(res_df) %>%
  # format numeric columns to 3 decimal places
  colformat_num(j = c("Estimate","Est.Error","l-95% CI","u-95% CI","Rhat","Bulk_ESS","Tail_ESS"),
                digits = 3) %>%
  # add a little zebra striping and autofit
  theme_zebra() %>%
  autofit() %>%
  # optional: bold the header row
  bold(i = 1, bold = TRUE)

# display
ft

```

@tbl-Model_outputs_Cyprus shows that the model's smooth term for sea‐surface temperature (SST) was estimated at --1.08 (95 % CI --2.22 to --0.10), indicating a clear, negative relationship whereby chlorophyll‐a declines markedly as SST increases. In contrast, the smooth effect of sea‐surface salinity was essentially zero (--0.003; 95 % CI --0.55 to 0.59), the N : P ratio exhibited a small positive coefficient (0.16; 95 % CI --0.92 to 1.23), and total nutrient load showed a moderate positive estimate (1.90; 95 % CI --0.23 to 4.05)---all of which had credible intervals overlapping zero. Likewise, mixed‐layer depth (--0.81; 95 % CI --3.15 to 0.71) and surface heat flux (0.36; 95 % CI --0.66 to 1.36) failed to demonstrate a statistically discernible influence on chlorophyll‐a. The Gamma‐family shape parameter was estimated at 98.91 (95 % CI 82.56 to 116.69). Taken together, only SST emerges as a significant driver of GeoMean_Chl a in this system.

```{r overall raw plot}
#| label: fig-Cyprus_Plots
#| fig-cap: "Posterior conditional (marginal) effects of the six environmental drivers retained in the Gamma-GAM for monthly geometric-mean chlorophyll-a (Chla) in the Gulf of Lyon. For each predictor, the solid coloured line is the posterior median of the expected Chla when that variable varies across its observed range while all other covariates are fixed at their means; the shaded band is the corresponding 95 % credible interval on the response scale. Axes are shown in original physical units: sea-surface temperature (°C), net heat flux (W m⁻²), molar N : P ratio, summed dissolved nutrients (mmol m⁻³), mixed-layer depth (m) and sea-surface salinity (ppt)."
#| echo: false
#| error: false
#| message: false
#| warning: false
#| fig-height: 10

plot_Cyprus
```

@fig-Cyprus_Plots displays the posterior mean smooths (solid lines) and their 95 % credible ribbons for each of the six covariates, expressed on the original physical scales. As sea‐surface temperature rises from roughly 17 °C to 28 °C, modeled Chl a falls from about 0.06 mg m⁻³ to 0.035 mg m⁻³, with the steepest decline between 17 and 22 °C before the curve levels off at warmer temperatures. Heat flux produces a classic U-shape: Chl a is near 0.042 mg m⁻³ under strong cooling (--300 W m⁻²), drops to \~0.040 mg m⁻³ around 0 W m⁻², then rises to \~0.045 mg m⁻³ by +200 W m⁻². The N : P ratio effect is more subtle, with Chl a increasing gradually from \~0.037 mg m⁻³ at an N : P of 25 to \~0.041 mg m⁻³ at N : P of 100 and a slight plateau of \~0.039 mg m⁻³ between ratios of 100--150. Total nutrient load drives a strong positive response, as Chl a climbs from \~0.028 mg m⁻³ at 0 mmol m⁻³ to \~0.04 mg m⁻³ by 50 mmol m⁻³, continues toward \~0.06 mg m⁻³ at 200 mmol m⁻³, and approaches \~0.07 mg m⁻³ at 300 mmol m⁻³. Mixed-layer depth showshave nearly no effect on chla between 0 and 75m of MLD, but a negative effect on chla can be seen when MLD is greater than 75m with chla going from 0.040 mg m⁻³at 75m to 0.030 mg m⁻³ at 175m . Finally, sea-surface salinity elicits virtually no change, with Chl a constrained between \~0.038 and \~0.040 mg m⁻³ over the 38.9--39.9 ppm range.
